Engineering a Runtime System for AQL

### 2016-03-07 paper/paper.tex
Almost complete stub of Chapter 3

Consider two spans $s_{1},s_{2}$. The main span predicates that may
be used for selection are the following:
\begin{lyxlist}{00.00.0000}
\item [{$s_{1}\preceq_{d}s_{2}$}] \noindent when $s_{1},s_{2}$ do not
overlap, $s_{1}$ precedes $s_{2}$ and there are less than $d$ characters
between them;
\item [{$s_{1}\simeq s_{2}$}] when the two spans overlap;
\item [{$s_{1}\subset s_{2}$}] when $s_{1}$ is strictly contained in
$s_{2}$;
\item [{$s_{1}=s_{2}$}] when $s_{1}$ equals $s_{2}$.
\end{lyxlist}

@@

\noindent Additionally, we might use $\gamma^{+}$ as a shortcut for
$\gamma\cdot\gamma^{\ast}$, and $\varSigma$ as an abbreviation of
$\sigma_{1}\vee...\vee\sigma_{n}$ (with a slight abuse of notation).
We also need the definition of a \emph{string relation}.
\begin{definition}
A n-ary string relation is a subset of $\left(\varSigma^{\ast}\right)^{n}$.
\end{definition}

@@

\noindent The spanner represented by a regex formula $\gamma$ may
be denoted as $\left\llbracket \gamma\right\rrbracket $. We have
that $\textrm{SVars}\left(\left\llbracket \gamma\right\rrbracket \right)=\textrm{SVars}\left(\gamma\right)$,
and the span relation $\left\llbracket \gamma\right\rrbracket \left(s\right)$
is the set $\left\{ \mu^{p}\mid p\textrm{ is a }\gamma-\textrm{parse for \ensuremath{s}}\right\} $,
where $\mu^{p}$ is a tuple defined by a parse $p$.

@@

\noindent The set of all possible accepting runs of a vstk-automaton
$A$ on a string $s$ is denoted as $\textrm{ARuns}\left(A,s\right)$.
The spanner represented by $A$ may be referred to as $\left\llbracket A\right\rrbracket $.
We have that $\textrm{SVars}\left(\left\llbracket A\right\rrbracket \right)=\textrm{SVars}\left(A\right)$,
and the span relation $\left\llbracket A\right\rrbracket \left(s\right)$
is the set $\left\{ \mu^{\rho}\mid\rho\in\textrm{ARuns}\left(A,s\right)\right\} $,
where $\mu^{\rho}$ is a tuple defined by a run $\rho$. In particular,
for every variable $x\in\textrm{SVars}\left(A\right)$, $\mu^{\rho}\left(x\right)$
is the span $\left[i_{b},i_{e}\right\rangle $, where:
\begin{itemize}
\item $c_{b}$=$\left(q_{b},\overrightarrow{v_{b}},Y_{b},i_{b}\right)$
is the unique configuration of $\rho$ where $x$ appears in the stack
for the fist time;
\item $c_{e}$=$\left(q_{e},\overrightarrow{v_{e}},Y_{e},i_{e}\right)$
is the unique configuration of $\rho$ where $x$ appears in the stack
for the last time.
\end{itemize}

@@

\noindent Given a string $s$ and a vset-automaton $A$, $\textrm{ARuns}\left(A,s\right)$and
$\left\llbracket A\right\rrbracket $ are defined as for vstk-automata.

\subsection{Algebras of Spanners}

@@

\subsubsection{Projection}

\subsubsection{Natural Join}

\subsubsection{String Selection}

>>>>

\noindent Consider two spans $s_{1},s_{2}$. The main span predicates
that may be used for selection are the following:
\begin{lyxlist}{00.00.0000}
\item [{$s_{1}\preceq_{d}s_{2}$}] \noindent when $s_{1},s_{2}$ do not
overlap, $s_{1}$ precedes $s_{2}$ and there are less than $d$ characters
between them;
\item [{$s_{1}\simeq s_{2}$}] when the two spans overlap;
\item [{$s_{1}\subset s_{2}$}] when $s_{1}$ is strictly contained in
$s_{2}$;
\item [{$s_{1}=s_{2}$}] when $s_{1}$ equals $s_{2}$.
\end{lyxlist}

@@

\noindent Additionally, we might use $\gamma^{+}$ as a shortcut for
$\gamma\cdot\gamma^{\ast}$, and $\varSigma$ as an abbreviation of
$\sigma_{1}\vee...\vee\sigma_{n}$ (with a slight abuse of notation).
The language $\mathcal{L}\left(\gamma\right)$of a regular expression
$\gamma$ is the set of strings $\mathbf{s}$ over $\varSigma$ that
are matched by that expression. A language $L$ is \emph{regular}
if $L=\mathcal{L}\left(\gamma\right)$ for some regular expression
$\gamma$. We also need the definition of a \emph{string relation}.
\begin{definition}
A n-ary string relation is a subset of $\left(\varSigma^{\ast}\right)^{n}$.
\end{definition}

\noindent An interesting class of string relations is that of recognizable
relations, denoted as $\textrm{REC}$.
\begin{definition}
Given a k-ary string relation $R$, $R$ is recognizable if it is
representable by a finite union the form:
\begin{equation}
\bigcup L_{1}\times...\times L_{k}
\end{equation}

where each $L$$_{i}$ is a regular language.
\end{definition}

@@

\noindent The spanner represented by a regex formula $\gamma$ may
be denoted as $\left\llbracket \gamma\right\rrbracket $. We have
that $\textrm{SVars}\left(\left\llbracket \gamma\right\rrbracket \right)=\textrm{SVars}\left(\gamma\right)$,
and the span relation $\left\llbracket \gamma\right\rrbracket \left(s\right)$
is the set $\left\{ \mu^{p}\mid p\textrm{ is a }\gamma-\textrm{parse for \ensuremath{s}}\right\} $,
where $\mu^{p}$ is a tuple defined by a parse $p$. The class of
regex formulas is referred to as $\textrm{RGX}$.

@@

\noindent The set of all possible accepting runs of a vstk-automaton
$A$ on a string $s$ is denoted as $\textrm{ARuns}\left(A,s\right)$.
The spanner represented by $A$ may be referred to as $\left\llbracket A\right\rrbracket $.
We have that $\textrm{SVars}\left(\left\llbracket A\right\rrbracket \right)=\textrm{SVars}\left(A\right)$,
and the span relation $\left\llbracket A\right\rrbracket \left(s\right)$
is the set $\left\{ \mu^{\rho}\mid\rho\in\textrm{ARuns}\left(A,s\right)\right\} $,
where $\mu^{\rho}$ is a tuple defined by a run $\rho$. In particular,
for every variable $x\in\textrm{SVars}\left(A\right)$, $\mu^{\rho}\left(x\right)$
is the span $\left[i_{b},i_{e}\right\rangle $, where:
\begin{itemize}
\item $c_{b}$=$\left(q_{b},\overrightarrow{v_{b}},Y_{b},i_{b}\right)$
is the unique configuration of $\rho$ where $x$ appears in the stack
for the fist time;
\item $c_{e}$=$\left(q_{e},\overrightarrow{v_{e}},Y_{e},i_{e}\right)$
is the unique configuration of $\rho$ where $x$ appears in the stack
for the last time.
\end{itemize}
\noindent The class of variable stack automata is called $\textrm{VA}_{\textrm{stk}}$.

@@

\noindent Given a string $s$ and a vset-automaton $A$, $\textrm{ARuns}\left(A,s\right)$and
$\left\llbracket A\right\rrbracket $ are defined as for vstk-automata.
The class of variable set automata is called $\textrm{VA}_{\textrm{stk}}$.

\subsection{Algebras of Spanners\label{sub:Algebras-of-Spanners}}

@@

Before giving the definition of the union operator, we need to introduce
the concept of \emph{union compatible spanners.}
\begin{definition}
Given two document spanner $P_{1}$ and $P_{2}$, they are union compatible
if and only if $\textrm{SVars}\left(P_{1}\right)=\textrm{SVars}\left(P_{2}\right)$.
\end{definition}

\noindent The definition of union of two spanners is as follows.
\begin{definition}
Given two union compatible document spanners $P_{1}$ and $P_{2}$,
their union is the spanner denoted as $P_{1}\cup P_{2}$, for which
we have that:
\begin{itemize}
\item $\textrm{SVars}\left(P_{1}\cup P_{2}\right)=\textrm{SVars}\left(P_{1}\right)$;
\item given a string $\mathbf{s}$, $\left(P_{1}\cup P_{2}\right)\left(\mathbf{s}\right)=P_{1}\left(\mathbf{s}\right)\cup P_{2}\left(\mathbf{s}\right)$.
\end{itemize}
\end{definition}

\subsubsection{Projection}
\begin{definition}
Given a document spanner $P$ and a set of span variables $Y\subseteq\textrm{SVars}\left(P\right)$,
the projection of $P$ over $Y$ is the spanner denoted as $\pi_{Y}P$,
satisfying $\textrm{SVars}\left(\pi_{Y}P\right)=Y$.
\end{definition}

\noindent As the name and the definition suggest, the projection of
a document spanner $P$ over a subset of its span variables $Y$ is
obtained by reducing the domain of each of the spanner's output $s-\textrm{tuples}$
to $Y$.

\subsubsection{Natural Join}
\begin{definition}
Given two document spanners $P_{1}$ and $P_{2}$, their natural join
is the spanner denoted as $P_{1}\bowtie P_{2}$, for which we have
that:
\begin{itemize}
\item $\textrm{SVars}\left(P_{1}\bowtie P_{2}\right)=\textrm{SVars}\left(P_{1}\right)\cup\textrm{SVars}\left(P_{2}\right)$;
\item given a string $\mathbf{s}$, $\left(P_{1}\bowtie P_{2}\right)\left(\mathbf{s}\right)$
consists of all the $\mathbf{s}-\textrm{tuples}$ $\mu$ agreeing
with some $\mu_{1}\in P_{1}\left(\mathbf{s}\right)$ and some $\mu_{2}\in P_{2}\left(\mathbf{s}\right)$.
Note that this implies that $\mu_{1},\mu_{2}$ agree on variables
that are common to $P_{1},P_{2}$: $\forall x\in\textrm{SVars}\left(P_{1}\right)\cap\textrm{SVars}\left(P_{2}\right)$,
$\mu_{1}\left(x\right)=\mu_{2}\left(x\right)$. 
\end{itemize}
\end{definition}

\subsubsection{String Selection}
\begin{definition}
Given a document spanner $P$ and a k-ary string relation $R$, the
string selection operation according to $R$ is denoted as $\varsigma^{R}$,
and is parametrized by $x_{1},...,x_{k}\in\textrm{SVars}\left(P\right)$.
We have that, given a string $\mathbf{s}$ and a spanner $P':=\varsigma_{x_{1},...,x_{k}}^{R}P$,
$P'\left(\mathbf{s}\right)$ consists of all the $\mathbf{s}-\textrm{tuples}$
$\mu$ such that $\left(\mathbf{s}_{\mu\left(x_{1}\right)},...,\mathbf{s}_{\mu\left(x_{k}\right)}\right)\in R$.
\end{definition}

\noindent In the remainder of this document, the only string selection
operator I consider is $\varsigma_{x,y}^{=}$ which, given a spanner
$P$, restricts $P\left(\mathbf{s}\right)$ to those $\mathbf{s}-\textrm{tuples}$
that satisfy $\mathbf{s}_{\mu\left(x\right)}=\mathbf{s}_{\mu\left(y\right)}$.

\medskip{}

\noindent I now introduce some additional notation. Given a generic
class of spanner representations $\textrm{SR}$, the set of all the
spannners that can be represented by $\textrm{SR}$ is denoted as
$\left\llbracket \textrm{SR}\right\rrbracket $. Formally, we have
that $\left\llbracket \textrm{SR}\right\rrbracket =\left\{ \left\llbracket r\right\rrbracket \mid\in\textrm{SR}\right\} $.
Let $O$ be a spanner algebra. I denote by $\textrm{SR}^{O}$ the
closure of $\textrm{SR}$ under $O$, that is: the class of spanner
represantations obtained by applying (compositions of) operators contained
in $O$ to the represantations in $\textrm{SR}$. The corresponding
set of spanners is referred to as $\left\llbracket \textrm{SR}^{O}\right\rrbracket $.
We are ready to start reasoning on the expressive power of the spanner
representations that we defined.
\begin{proposition}
The following hold:
\begin{enumerate}
\item every document spanner represented in the classes $\textrm{RGX}$
and $\textrm{VA}_{\textrm{stk}}$ is hierarchical, that is: $\left\llbracket \textrm{\textrm{RGX}}\right\rrbracket ,\left\llbracket \textrm{\ensuremath{\textrm{VA}_{\textrm{stk}}}}\right\rrbracket \subseteq\mathbf{HS}$;
\item there exist some spanners represented in $\textrm{VA}_{\textrm{set}}$
that are not hierarchical: $\left\llbracket \textrm{\ensuremath{\textrm{VA}_{set}}}\right\rrbracket \nsubseteq\mathbf{HS}$;
\item the operators $\cup$, $\pi$, $\varsigma^{R}$ preserve the property
of being hierarchical, while $\bowtie$ does not, thus we have that:

\begin{enumerate}
\item given a class of spanner representations $\textrm{SR}$, $\left\llbracket \textrm{SR}\right\rrbracket \subseteq\mathbf{HS}\Rightarrow\left\llbracket \textrm{SR}^{\left\{ \cup,\pi,\varsigma^{R}\right\} }\right\rrbracket \subseteq\mathbf{HS};$
\item there exist two hierarchical spanners $P_{1},P_{2}$ such that $P_{1}\bowtie P_{2}\notin\mathbf{HS}$.
\end{enumerate}
\end{enumerate}
\end{proposition}

\noindent In the next Subsection, I discuss the class of \emph{regular
spanners}, which plays a central role in the discussion of the class
of spanners that constitute the core of AQL.

\subsection{Regular Spanners}

A regular spanner is defined as follows.
\begin{definition}
A spanner is regular if it can be defined by a vset-automaton.
\end{definition}

\noindent Let us now see how regular spanners are related to the other
basic spanner classes. A preliminary result is that regex formulas
and vstk-automata have the same expressive power.
\begin{theorem}
\label{.rgx=00003Dvstk}$\left\llbracket \textrm{\textrm{RGX}}\right\rrbracket =\left\llbracket \textrm{\ensuremath{\textrm{VA}_{\textrm{stk}}}}\right\rrbracket $.
\end{theorem}

\noindent It turns out that the spanners expressed by representations
in $\textrm{\ensuremath{\textrm{VA}_{\textrm{stk}}}}$($\textrm{\textrm{RGX}}$)
are exactly those that are both regular and hierarchical.
\begin{theorem}
$\left\llbracket \textrm{\ensuremath{\textrm{VA}_{\textrm{stk}}}}\right\rrbracket =\left\llbracket \textrm{\ensuremath{\textrm{VA}_{\textrm{set}}}}\right\rrbracket \cap\mathbf{HS}$.
\end{theorem}

\noindent For what concerns the algebraic operators I presented in
Subsection \ref{sub:Algebras-of-Spanners}, it can be shown that union,
projection and join don't increase the expressive power of regular
spanners.
\begin{theorem}
\label{.setext=00003Dset}$\left\llbracket \textrm{\ensuremath{\textrm{VA}_{\textrm{set}}}}^{\left\{ \cup,\pi,\bowtie\right\} }\right\rrbracket =\left\llbracket \textrm{\ensuremath{\textrm{VA}_{\textrm{set}}}}\right\rrbracket $.
\end{theorem}

\noindent On the other hand, the same operators make the spanners
represented in $\textrm{\ensuremath{\textrm{VA}_{\textrm{stk}}}}$
equivalent to regular spanners.
\begin{theorem}
\label{.stkext=00003Dset}$\left\llbracket \textrm{\ensuremath{\textrm{VA}_{\textrm{stk}}}}^{\left\{ \cup,\pi,\bowtie\right\} }\right\rrbracket =\left\llbracket \textrm{\ensuremath{\textrm{VA}_{\textrm{set}}}}\right\rrbracket $.
\end{theorem}

\noindent Finally, I look at which string relations can be simulated
by regular spanners. Let us start by defining the concept of \emph{selectable
string relation}.
\begin{definition}
Given a string relation $R$ and a class of spanners $C$, $R$ is
selectable by $C$ if for every document spanner $P\in C$ and for
every $\overrightarrow{x}=x_{1},...,x_{k}$, $x_{i}\in\textrm{SVars}\left(P\right)$,
we have that $\varsigma_{\overrightarrow{x}}^{R}P\in C$.
\end{definition}

\noindent Next, I introduce the concept of \emph{restricted universal
spanner.}
\begin{definition}
Given a string relation $R$ and a sequence of variables$\overrightarrow{x}=x_{1},...,x_{k}$
with their corrsesponding set $X=\left\{ x_{1},...,x_{k}\right\} $,
the $R-\textrm{restricted}$ universal spanner over $\overrightarrow{x}$
is $\Upsilon_{\overrightarrow{x}}^{R}:=\varsigma_{\overrightarrow{x}}^{R}\Upsilon_{X}$.
\end{definition}

\noindent Selectability of a string relation $R$ by a class of spanners
$C$ corresponds to the presence in $C$ of all the possible $R-\textrm{restricted}$
universal spanners, under some conditions.
\begin{proposition}
Given a string relation $R$ and a class of spanners $C$ containing
all the possible universal spanners and closed under natural join,
$R$ is selectable by $C$ if and only if, for every $\overrightarrow{x}=x_{1},...,x_{k}\in\textrm{SVars}^{k}$,
$\Upsilon_{\overrightarrow{x}}^{R}\in C$.
\end{proposition}

\noindent In the case of regular spanners, the class of string relations
that they can select is exactly $\textrm{REC}$.
\begin{theorem}
The class of string relations selectable by $\left\llbracket \textrm{\ensuremath{\textrm{VA}_{\textrm{set}}}}\right\rrbracket $
is $\textrm{REC}$.
\end{theorem}

\noindent The relation ``$=$'' is not in $\textrm{REC}$, thus
it is not selectable by regular spanners, nonetheless it is important
for selection predicates in AQL, which is why regular spanners are
not enough to model its core. In the following Subsection, by using
the notions and results presented so far, I describe the spanners
that are identified in REF as the core of AQL: \emph{core spanners}.

\subsection{Core Spanners}

An expression in the core of AQL belongs to $\textrm{RGX}^{\left\{ \cup,\pi,\bowtie,\varsigma^{=}\right\} }$.
Consequently, a core spanner is defined as follows.
\begin{definition}
A core spanner is a document spanner belonging to $\left\llbracket \textrm{RGX}^{\left\{ \cup,\pi,\bowtie,\varsigma^{=}\right\} }\right\rrbracket $.
\end{definition}

\noindent Thanks to Theorems \ref{.rgx=00003Dvstk}, \ref{.setext=00003Dset}
and \ref{.stkext=00003Dset} we can easily see that $\left\llbracket \textrm{RGX}^{\left\{ \cup,\pi,\bowtie,\varsigma^{=}\right\} }\right\rrbracket =\left\llbracket \textrm{\ensuremath{\textrm{VA}_{\textrm{stk}}}}^{\left\{ \cup,\pi,\bowtie,\varsigma^{=}\right\} }\right\rrbracket =\left\llbracket \textrm{\ensuremath{\textrm{VA}_{\textrm{set}}}}^{\left\{ \cup,\pi,\bowtie,\varsigma^{=}\right\} }\right\rrbracket $.
This shows that core spanners can be reduced to regular spanners extended
with the algebra $\left\{ \cup,\pi,\bowtie,\varsigma^{=}\right\} $.
But the following lemma tells us an algebra with fewer operators is
also sufficient.
\begin{lemma}
$\left\llbracket \textrm{\ensuremath{\textrm{VA}_{\textrm{set}}}}^{\left\{ \cup,\pi,\bowtie,\varsigma^{=}\right\} }\right\rrbracket =\left\llbracket \textrm{\ensuremath{\textrm{VA}_{\textrm{set}}}}^{\left\{ \pi,\varsigma^{=}\right\} }\right\rrbracket $.
\end{lemma}

\noindent Another lemma, known as the \emph{core simplification lemma},
gives a even more simple way of representing core spanners.
\begin{lemma}
$\mathbf{(Core\,Simplification\,Lemma)}$ Every core spanner can be
defined by an expression of the form

\begin{equation}
\pi_{V}SA
\end{equation}

\noindent where:
\begin{itemize}
\item $A$ is a vset-automaton;
\item $V\subseteq\textrm{SVars}\left(A\right)$;
\item $S$ is a sequence of string selections $\varsigma_{x,y}^{=}$, for
$x,y\in\textrm{SVars}\left(A\right)$.
\end{itemize}
\end{lemma}

\noindent Now I look at which string relations can be simulated by
core spanners. In the next definition, I define three string relations
of relevance.
\begin{definition}
Given two strings $\mathbf{s},\mathbf{t}\in\varSigma^{\ast}$:
\begin{itemize}
\item $\mathbf{s}\sqsubseteq\mathbf{t}$ if $\mathbf{s}$ is a (consecutive)
substring of t (i.e. $\mathbf{s}=\mathbf{t}_{\left[i,j\right\rangle }$);
\item $\mathbf{s}\sqsubseteq_{\textrm{prf}}\mathbf{t}$ if $\mathbf{s}$
is a prefix of t (i.e. $\mathbf{s}=\mathbf{t}_{\left[1,j\right\rangle }$);
\item $\mathbf{s}\sqsubseteq_{\textrm{sfx}}\mathbf{t}$if $\mathbf{s}$
is a suffix of t (i.e. $\mathbf{s}=\mathbf{t}_{\left[1,\left|\mathbf{t}\right|+1\right\rangle }$).
\end{itemize}
\end{definition}

\begin{proposition}
\noindent All the string relations in $\textrm{REC}$, $\mathbf{\sqsubseteq}$,
$\sqsubseteq_{\textrm{prf}}$ and $\sqsubseteq_{\textrm{sfx}}$ are
selectable by the core spanners.
\end{proposition}

\subsection{Difference}

The difference operator is defined as follows.
\begin{definition}
Given two union compatible document spanners $P_{1}$ and $P_{2}$,
their difference is the spanner $P_{1}\setminus P_{2}$, for which
we have:
\begin{itemize}
\item $\textrm{SVars}\left(P_{1}\setminus P_{2}\right)=\textrm{SVars}\left(P_{1}\right)$;
\item given a string $\mathbf{s}$, $\left(P_{1}\setminus P_{2}\right)\left(\mathbf{s}\right)=\left(P_{1}\right)\setminus\left(P_{2}\right)\left(\mathbf{s}\right)$.
\end{itemize}
\end{definition}

\noindent It can be shown that regular spanners are closed under difference.
\begin{theorem}
\label{.setdiff=00003Dset}$\left\llbracket \textrm{\ensuremath{\textrm{VA}_{\textrm{set}}}}^{\left\{ \setminus\right\} }\right\rrbracket =\left\llbracket \textrm{\ensuremath{\textrm{VA}_{\textrm{set}}}}\right\rrbracket $.
\end{theorem}

\noindent Despite the result of of Theorem \ref{.setdiff=00003Dset},
core spanners are not closed under difference, which is why this operator
has not been considered so far.
\begin{theorem}
$\left\llbracket \textrm{RGX}^{\left\{ \cup,\pi,\bowtie,\varsigma^{=}\right\} }\right\rrbracket \subsetneq\left\llbracket \textrm{RGX}^{\left\{ \cup,\pi,\bowtie,\varsigma^{=},\setminus\right\} }\right\rrbracket $.
\end{theorem}
